<style>@import url(style.css);</style>
<small>[Introduction to Data Analysis](index.html)</small>

# 1.1. Installing R

[R](http://www.r-project.org/) is a software that uses a specific language to manipulate statistical objects. Researchers and statisticians are [increasingly](http://r4stats.com/articles/popularity/) adopting R as their software of choice, and the R user base goes well beyond academia. Many [R User Groups](http://www.r-bloggers.com/RUG/) exist on a [worldwide scale](http://blog.revolutionanalytics.com/local-r-groups.html).

Some people *really* like R. This very enthusiastic user is [Anthony Damico](https://github.com/ajdamico), a [prolific](http://www.ajdamico.com/p/work.html) and [proficient](http://www.asdfree.com/) analyst of U.S. survey data working at the Kaiser Family Foundation. His video mentions other programming and statistical languages:

<iframe src="http://player.vimeo.com/video/52999628?badge=0" width="800" height="600" frameborder="0" webkitAllowFullScreen mozallowfullscreen allowFullScreen></iframe>

The R language has its roots in the S language developed by AT&T, which also developed the C language. It is not the only language available for statistical analysis: there are [many others](http://brenocon.com/blog/2009/02/comparison-of-data-analysis-packages-r-matlab-scipy-excel-sas-spss-stata/), and you might have also heard about SAS, SPSS or Stata, or even Matlab and Python.

R is as free as it is powerful, but it comes with [some drawbacks](http://badhessian.org/seven-reasons-to-use-r-for-social-network-analysis-and-three-reasons-against/). Its major disadvantage is that its learning curve is pretty steep. Do not worry: we will review as many examples as we need to make it work, and we will give you code to help creating your own.

## Installation

Now, download  and install the latest version of R. You will need admin privileges on your computer (i.e. the user login and password) to do so. R is free and available for all platforms. It does not take a lot of memory to install and use.

The R development team maintains [a website](http://www.r-project.org/), but you will download R from a different source based in the United States. The exact download link depends on your operating system:

- __For Mac OS X__, [download from this link](http://cran.us.r-project.org/bin/macosx/), double-click it, say "OK" to all installation steps. You can also watch [this video tutorial](https://www.youtube.com/watch?v=Icawuhf0Yqo) by Roger Peng for help, or [this video tutorial](https://www.youtube.com/watch?v=WJDrYUqNrHg&list=PL8BE0E317807A9A21) by Gordon Davis, which also covers RStudio.
- __For Windows__, [download from this link](http://cran.us.r-project.org/bin/windows/base/), which also has a few help pages. [This video tutorial](https://www.youtube.com/watch?v=mfGFv-iB724) by Roger Peng and [this video tutorial](http://www.screenr.com/kzT8) by Anthony Damico can help with the installation. You might also consider [using a few custom options](https://www.youtube.com/watch?v=Lc2sgDTzrV8) when running R on Windows.

Do not worry if you do not manage to install R (or RStudio, the next software that you will have to install) on your own. Just download the program to your hard drive, and we will guide you through installation in class.

## Commands

Open R and locate the blinking cursor preceded by a `>` at the bottom of the R Console window. This is where you will type commands and read their results. You might be already familiar with that process if you have ever used a Command Line Interface (CLI) before.

Type the following lines in R and press Enter to execute, or "run", their commands. You should skip the lines that start with a `#` (hash) symbol and show in a different color than the rest of the code: these lines are comments, which R will ignore if you try to run them.

```{r very-first-commands, results='hide'}
# A string of characters.
"Hello R World!"
# A numeric result.
1 + 2
# A logical statement.
1 / (2 + 3) == .2
# A vector of integers.
1:3
```

If you get an error message in red ink at any point, it means that you have run into a syntax error: press &uarr; (UpArrow) to go back to your last command, check your typing against the original shown here, correct it, and press Enter to try it again.

The next section tells you a few more things about R syntax.

## Syntax

You might have run into errors in the example above if you typed anything else than the code provided. That is because R, just like every other programming language, requires that you follow a precise syntax. You might have run into these errors in particular:

- __Forgetting quotes around text.__ R requires that you enclose text into single or double quotes, as to specify that it is text that you are typing, and not the name of an object. Try typing `hello` without quotes to see what an error message looks like in R.
- __Double equal signs in logical expressions.__ One of the commands in the example above is a logical statement and therefore requires that you type _two_ equal signs instead of just one as you would in common mathematics. R uses the single equal sign for assignment.
- __Typing UPPERCASE instead of lowercase.__ Note, finally, that R is case sensitive: lowercase and UPPERCASE letters are interpreted differently, and therefore cannot be used interchangeably. Lowercase is the common default, but UPPERCASE is also used from time to time.

Try the example below to see how changing the case can affect a given command:

```{r letters}
# A vector of lowercase letters.
letters[1:5]
# A vector of UPPERCASE letters.
LETTERS[1:5]
```

One last element of R syntax that you will have to get familiar with is the use of punctuation. Brackets and commas, in particular, are put to intensive use in R. The examples below show some of their common usage in R:

```{r}
# A sequence of floating point numbers.
seq(from = 1, to = 3, by = .5)
# A function with a logical argument.
order(1:3, decreasing = TRUE)
```

Whitespace is technically ignored by R, so leaving space after a comma, for example, is not important for the successful execution of a command. Forgetting a comma or a closing bracket, however, will end up in a syntax error.

The next section tells you a bit more about objects and assignment, which were both mentioned in the explanations above. We will come back to brackets and commas later, when we study functions. 

## Assignment

Type each command below in their order of appearance. The commands will _not_ produce any (immediate) result, which is normal: just carry on. If you are uncomfortable with using the `<-` symbol, you can type `=` instead: the two symbols are equivalent in R.

```{r assign}
# Create an object called x.
x <- "Hello"
# Create an object called y.
y <- "World"
```

This example shows you how to assign a value to an object in R. The basic operator `<-` assigns the values `"Hello"` and `"World"` to the objects `x` and `y`. We will now bind these objects together into the object `z` with the `c()` command.

```{r combine}
# Combine x and y into a vector called z.
z <- c(x, y)
```

The result can be shown with the `print()` function, or just by typing the name of the object.

```{r print-results}
# Print the object z on screen.
print(z)
# Just type its name do do the same.
z
```

R syntax is very much like German or Latin syntax: a bit counter-intuitive at first, but flawlessly logical. It takes a lot of practice to feel at ease with it, but it will never fail you. Just work through every example in the next pages to learn it step by step.

## Exit

You can quit R like any other application or by typing `q()` from the command line interface. You might be asked whether you want to save the R session that you started by creating the `x`, `y` and `z` objects in your Workspace (more on that later). You do not need to do so.

R can do a lot of things, including [waiting for you to make coffee](http://vimeo.com/43305640). A big drawback of R, however, is its barebones interface. We will fix that by installing a software that "pilots" R through better menus and windows.

> __Next__: [Installing RStudio](012_rstudio.html).
