<style>@import url(style.css);</style>
[Introduction to Data Analysis](index.html "Course index")

# 2. Objects

<!-- some reorganization needed from "let's review syntax" to the end -->

One of the first things that you want to learn in R is to manipulate objects. In R, an object can be many things: a number, a piece of text, a function, a series of values, a dataset… Let's start with the very fundamentals.

R can manipulate numbers. You certainly suspect that R can compute things pretty much like a scientific calculator or like [Wolfram Alpha](https://www.wolframalpha.com/). For that matter, the following are valid mathematical expressions in R, and the results they return are numerical in nature:

```{r calc, results='hide'}
# Addition.
1 + 2
# Brackets.
(1 + 2) / 3
# Powers.
7^2
# Infinity.
1 - Inf
# And so on. These objects are...
class(7^2)
```

These commands have created objects of a particular class: numeric objects. Train yourself by computing, for instance, $3^3 - 4$ and $e^2$. If you need help to find out how to code an exponential in R, just look for help with `?exp`.

From mathematics, you also know about logical statements, which can only be true or not-true (false). You can ask R to evaluate a given expression by submitting it as a logical expression, and R will respond with either `TRUE` or `FALSE`:

```{r logical, results='hide'}
# A simple logical test with the natural logarithm.
log(1) > 0
# Guess why this result differs from the one above.
log(1) >= 0
# Equality requires TWO equal signs.
log(1) == 0
# Negation requires the "!" symbol.
log(1) != 1
# A more complex example using scientific notation.
1 + log10(1e7) == 8 * exp(1)^0
# And so on. These objects are...
class(1 > 1)
```

These commands have created objects of another class: logical objects.

You might also remember manipulating text, or "strings", from our first session. There are several ways to have R "paste" and concatenate strings together, which is useful as soon as you are manipulating data like names or dates. Here are some examples:

```{r character}
# The typical "hello world" text.
"Hello R World!"
# Now "pasted" from its elements.
paste("Hello", "R", "World", "!")
# Now through raw concatenation.
cat("Hello", "R", "World", "!")
# And so on. These objects are...
class("some text")
```

R can also handle complex objects that hold more than one value. The example below defines a simple sequence of integer values $(1, 2, 3)$, computes its sum and product, and then compares them with each other:

```{r calc-vector, results='hide'}
# Create a sequence of integers.
1:3
# Compute the sum of the sequence.
sum(1:3)
# Compute the product of the sequence.
prod(1:3)
# Compare the sum and product.
sum(1:3) == prod(1:3)
```

The sequence defined above is called a numeric vector. It shows up in your Workspace, along with an indication that it contains integer values and has a dimension of 3. The rest of this session will train you at manipulating such objects.

Before we go further into manipulating objects, here's a quick tour of R syntax. You will inevitably make mistakes when writing R commands, which is fine -- nothing (bad) happens when you do. To correct a mistake, simply press `UpArrow` (&uarr;) to go back into your last console input and fix it.

```{r broken-hello-world, eval=FALSE, tidy=FALSE}
# This will break. Try it.
print(Hello World!)
# Press UpArrow, select text, add double quotes.
# This will now run fine.
print("Hello World")
```

R syntax can be a bit strange. For some (probably good) reason, the `install.packages` command takes quotes, whereas the `library` command does not.

```{r quotes-strangeness, eval=FALSE}
# This needs quotes.
install.packages("ggplot2")
# This does not.
library(ggplot2)
# But it can.
library("ggplot2")
# There's also another (less noisy) way to load packages.
require(RCurl)
# Thankfully, it also accepts quotes.
require("RCurl")
# Oh, and single quotes can work too...
library('ggplot2')
# ... And that covers the whole spectrum of quotes strangeness.
require('RCurl')
```

Note that brackets are omnipresent in R syntax: every function (or command) will require them. The bits written inside the brackets and separated by commas are called arguments. The `cat` command, for instance, accepts any number of arguments.

```{r cat}
# Print some text, the full date and some text again, with no separators.
cat("Hello World! Today is ", date(), ".", sep = "")
```

A final thing: it is good practice to write code in a certain style. For example, R is insensitive to extra blank space, but it is good practice to use blank space only when it helps to align elements. The [Google R Style Guide](https://google-styleguide.googlecode.com/svn/trunk/google-r-style.html) is a good starting point to learn a few common conventions.

Let's now turn to manipulating simple things in R.

The syntax of each command requires a bit of memorization, and lots of practice. Nobody types functional code in R without going through cycles of trial-and-error with it. Programming requires some discipline, some logic and a lot of patience.

> __Next__: [Vectors and matrices](021_vectors.html).
