<style>@import url(style.css);</style>
[Introduction to Data Analysis](index.html "Course index")

# 2.2. Variables and factors

In R, everything is an object. The list of objects in memory `ls()` is an object itself, as shown in the example below, which lists all objects in the current workspace and wipes them from memory.
  
```{r clear-workspace, eval=FALSE}
# List workspace objects.
ls()
# Erase entire workspace.
rm(list = ls())
```

Let's use some real, anonymized data from Autumn 2012. These are the grades from my three mathematics classes in first year. I have removed any student identification, so you have to trust me that these are the real grades (and yes, some grades range above 20/20!).  The `downloader` package provides a handy command to download the data from the course repository, so we install and load it first.

```{r grades-import}
# Install downloader package.
if(!"downloader" %in% installed.packages()[, 1])
  install.packages("downloader")
# Load package.
require(downloader)
# Target file.
file = "data/grades.2012.txt"
# Download the data if needed.
if (!file.exists(file)) {
  # Locate the data.
  url = "https://raw.github.com/briatte/ida/master/data/grades.txt"
  # Download the data.
  download(url, file, mode = "wb")
}
# Load the data.
grades <- read.table(file, header = TRUE)
# Check result.
head(grades)
```

Let's now use a package to create fake names for the students. We again need to install and load the package first: in later sessions, we will use a standard code block to install-and-load packages.

```{r grades-with-random-names}
# Install randomNames package. Remember that R is case-sensitive.
if(!"randomNames" %in% installed.packages()[, 1])
  install.packages("randomNames")
# Load package.
require(randomNames)
# How many rows of data do we have?
(count = nrow(grades))
# Let's generate that many random names.
names <- randomNames(count)
# Let's finally stick them to the matrix.
grades <- cbind(grades, names)
# Check result.
head(grades)
```

## Data frames

Let's show a final type of object: the data frame.

```{r df}
# Convert to data frame.
grades <- as.data.frame(grades)
# Check result.
head(grades)
# Check structure of a data frame.
str(grades)
```

Data frames are very malleable objects: we can rearrange the variables easily with commands like `melt` from the `reshape` package.

```{r melt, message=FALSE}
# Install and load reshape package.
if(!"reshape" %in% installed.packages()[, 1])
  install.packages("reshape")
# Load package.
require(reshape)
# Reshape data from 'wide' (lots of columns) to 'long' (lots of rows).
grades <- melt(grades, id.vars = "names")
# Check result to show how each grade is now held on a separate row.
head(grades[order(grades$names), ])
```

Let's finish with a few plots.

```{r grades-plots, tidy = FALSE}
# Install and load ggplot2 package.
if(!"ggplot2" %in% installed.packages()[, 1])
  install.packages("ggplot2")
# Load package.
require(ggplot2)
# Plot all three exams.
qplot(data = grades, x = value, 
      group = variable, 
      geom = "density")
# Add color and transparency.
qplot(data = grades, x = value, 
      color = variable, 
      fill = variable, 
      alpha = I(.3), geom = "density")
```

Now use the code on this page to:

1. Download [this data extract from the U.S. National Health Interview Survey 2005][nhis-data]. Use `RCurl` as shown above. Call the data `nhis`.

[nhis-data]: https://raw.github.com/briatte/ida/master/data/nhis.2005.txt

```{r nhis-data-import, include=FALSE}
file = "data/nhis.2005.txt"
# Download the data if needed.
if (!file.exists(file)) {
  # Locate the data.
  url = "https://raw.github.com/briatte/ida/master/data/nhis.2005.txt"
  # Download the data.
  download(url, file, mode = "wb")
}
# Load the data.
nhis <- read.table(file, header = TRUE)
# Check result.
head(nhis)
```

2. Create an object called `bmi` that corresponds to the Body Mass Index from the `height` and `weight` columns of the `nhis` object. Use the U.S. formula since the data use inches and pounds. Bind the `bmi` object to the `nhis` object.

```{r nhis-bmi, include=FALSE}
# Compute BMI vectore.
bmi <- 703 * nhis[ ,4] / nhis[ ,3]^2
# Bind to matrix.
nhis <- cbind(nhis, bmi)
```

3. Plot the results using `qplot(data = ..., x = ..., geom = "density")`.

```{r nhis-qplot-1, include=FALSE}
# The plot.
qplot(data = nhis, x = bmi, geom = "density")
```

4. Bonus question: explore how `ggplot2` works and produce plots with the `x` and `y` variables. Guess what they stand for.

```{r nhis-qplot-2, include=FALSE}
# Another plot. Variable x stands for gender.
qplot(data = nhis, x = bmi, group = x, color = factor(x), geom = "density")
# One more plot. Variable y stands for for racial group.
qplot(data = nhis, x = bmi, group = x, color = factor(x), geom = "density") +
  facet_grid(y ~ .)
```

> __Next__: [Practice](023_practice.html).
