<style>@import url(style.css);</style>
<small>[Introduction to Data Analysis](index.html)</small>

# 3. Basic math operators

We saw a bit of calculus, logic and vector/matrix manipulation last week. This week, we thread on the same topics with slightly more advanced operators. Here, for instance, is the modulus operator in R:

```{r modulus}
4 %/% 3
6 %/% 3
```

Many matrix operators are also available: you might want to go back to the [cheat sheet](https://gastonsanchez.wordpress.com/2012/08/30/cheat-sheet-of-matrix-operations-in-r/) previously mentioned. Here are some basic manipulations on matrixes built out of random integers.

```{r matrix-ops}
# Create a random 3 x 5 matrix.
A <- matrix(as.integer(10*runif(30)), nrow = 3, ncol = 5)
# Check result.
A
# Create a random 2 x 2 (square) matrix.
B <- matrix(as.integer(10*runif(16)), nrow = 2, ncol = 2)
# Check result.
B
# Create another one.
C <- matrix(as.integer(10*runif(16)), nrow = 2, ncol = 2)
# Check result.
C
# Now a basic manipulation: scalar multiplication.
2*A
# Another one: extract the diagonal.
diag(B)
# Last one: matrix transposition.
t(C)
```

As an exercise, explain the result of a square matrix product.

```{r matrix-product}
# Square matrix multiplication.
B %*% C
```

> __Next__: [Functions](31_functions.html).
