<style>@import url(style.css);</style>
[Introduction to Data Analysis](index.html "Course index")

# 3. Functions

We saw a bit of calculus, logic and vector/matrix manipulation last week. This week, we thread on the same topics with slightly more advanced operators. Here, for instance, is the modulus operator in R:

```{r modulus}
4 %/% 3
6 %/% 3
```

Many matrix operators are also available: you might want to go back to the [cheat sheet](https://gastonsanchez.wordpress.com/2012/08/30/cheat-sheet-of-matrix-operations-in-r/) previously mentioned. Here are some basic manipulations on matrixes built out of random integers.

```{r matrix-ops}
# Create a random 3 x 5 matrix.
A <- matrix(as.integer(10*runif(30)), nrow = 3, ncol = 5)
# Check result.
A
# Create a random 2 x 2 (square) matrix.
B <- matrix(as.integer(10*runif(16)), nrow = 2, ncol = 2)
# Check result.
B
# Create another one.
C <- matrix(as.integer(10*runif(16)), nrow = 2, ncol = 2)
# Check result.
C
# Now a basic manipulation: scalar multiplication.
2*A
# Another one: extract the diagonal.
diag(B)
# Last one: matrix transposition.
t(C)
```

As an exercise, explain the result of a square matrix product.

```{r matrix-product}
# Square matrix multiplication.
B %*% C
```

## Functional programming

Suppose that you find yourself doing the same computation over and over again. You do not want to type the whole calculations every time you need: you want to program a function that does the job for you.

Take a basic example: the `sum` command adds a vector of numbers together.

```{r sum}
# Create a vector of 99 random [0,1] values.
x <- runif(9)
# Check result: show first 5 values.
head(x)
# Add them up.
sum(x)
```

```{r add-function}
# Define function
add <- function(x,y) {
  x + y
}
# Example.
sum(2,4)
```

Functional programming focuses on writing functions. Statistical computing courses like [the one taught by Cosma Shalizi](http://www.stat.cmu.edu/~cshalizi/statcomp/) contains tons of interesting examples of such functions. A simple one is shown below to illustrate the principle of recursion.

```{r crs-factorial}
# Calculate a factorial
# Input: a number (n)
# Output: the factorial of n
# Presumes: n is a single positive integer
my.factorial <- function(n) {
  if (n == 1) {  # Base case
    return(1)
  } else {  # Recursion
    return(n*my.factorial(n-1))
  }
}
```

## Plotting functions

This course might lead you to write simple functions, but for the moment, let's focus simply on plotting them with the `ggplot2` package, which makes it fairly easy. The example below shows the basic function $y = x$.

```{r fun-identity}
qplot(c(0,2), stat="function", fun=identity, geom="point")
```

Now for the most math-oriented, here's a sine function for your enjoyment.

```{r fun-sin}
qplot(c(-10,10), stat="function", fun=sin, geom="line")
```

And finally, a function that will be more useful to us: the exponential.

```{r fun-exp}
qplot(c(-10:10), stat="function", fun=exp, geom=c("line","point"))
```

> __Next__: [Conditionals](031_conditionals.html).
