<style>@import url(style.css);</style>
[Introduction to Data Analysis](index.html "Course index")

# 3.2. Iteration

<!-- it's a mess -->

## `for` loops

R understands `for` and `while` loops like many other programming languages. The idea is the same for both operators: allow to repeat an operation a certain number of times. The first example is as trivial as it gets.

```{r for}
# Set the counter.
n <- 3
# Loop n times.
for(i in 1:n) {
  cat("Hello!", i, "\n")
}
```

The next example shows how to build a sequence of integers through a `for` loop. Explain the output.

```{r r-for-vector}
# Set the counter.
n <- 9
# Create an empty vector.
sequence <- NULL
# Loop over 9 values.
for(i in 1:n) {
  # Add i to the sequence.
  sequence <- c(sequence, i)
  # Mean of the sequence.
  print(mean(sequence))
}
# Show final result.
sequence
```

## `while` loops

A `while` loop operates [almost on the same level](https://en.wikipedia.org/wiki/While_loop). It iterates *until* a condition becomes true, or, if you prefer, *while* a condition is not *yet* true. Be careful: if the condition never becomes true, your loop will iterate [indefinitely](https://en.wikipedia.org/wiki/Infinite_loop).

```{r while}
# Set the counter.
i <- 5
# Decrease i by an increment of 1 until it reaches 0.
while(i > 0) {
  print(i <- i - 1)
}
```

## Extensions

You can do amusing things with loops, like [solving number games](https://xianblog.wordpress.com/2013/02/22/le-monde-puzzle-809/). However, as the author of the example writes, it is probably more fun to find the math trick behnd the puzzle than to 'brute-force' the game by trying out all possible numbers through a loop.

Loops have this 'brutal' aspect that actually make them quite unpopular, and they can become memory-hungry when they iterate a lot. R instead privileges the vectorization of iterated functions through [`apply` procedures](http://badhessian.org/loops-matrices-and-apply-functions/) that can often do what you might want to do with a loop.

There is another way to deal more efficiently with loops, which is to run them over several computer processors. If you are running a dual-core computer, you can run loops as parallel computing jobs with the [`foreach` package](http://cran.r-project.org/web/packages/foreach/index.html).

> __Next__: [Practice](033_practice.html).
