<style>@import url(style.css);</style>
[Introduction to Data Analysis](index.html "Course index")

# 5.3. Practice

<p class="info"><strong>Instructions:</strong> this week's exercise is called <code><a href="5_confidence.R">5_confidence.R</a></code>. Download or copy-paste that file into a new R script, then open it and run it with the working directory set as the <code>IDA</code> folder. If you download the script, make sure that your browser preserved its <code>.R</code> file extension.</p>

```{r run-exercise, include = FALSE, results = 'hide'}
source("code/5_confidence.R")
```

The code shows a selection of base PCA functions, followed by short examples showing how to produce [simple biplots][gh-ggbiplot] and [dendogram plots][gs-ggdendro] for PCA results obtained with the `FactoMineR` and `ClustOfVar` packages, which support more situations than the base PCA functions.

[gh-ggbiplot]: https://github.com/vqv/ggbiplot
[gs-ggdendro]: https://gastonsanchez.wordpress.com/2012/10/03/7-ways-to-plot-dendrograms-in-r/

```{r ggbiplot-auto, echo = FALSE}
# ggbiplot
ggbiplot(qog.prcomp,
         labels = rownames(qog),
         groups = countrycode(rownames(qog), "iso3c", "continent"),
         ellipse = TRUE) + 
  coord_equal(xlim = c(-3, 3), ylim = c(-3, 3))
```

> __Next week__: [Distributions](060_distributions.html).
