<style>@import url(style.css);</style>
<small>[Introduction to Data Analysis](index.html)</small>

# 8.2. Ordinary least squares

We continue with the QOG example, switching to the fertility rate.

<!-- For starters, let's try to predict some things.

- 
- Let's continue with the [2012 U.S. presidential election](http://www.douglas-hibbs.com), using a model that was [highly effective](http://lanekenworthy.net/2008/02/03/bread-peace-and-the-2008-election/) for past elections. -->

Some of the general ideas that emerge at that stage are:

1. The logic of "ordinary least squares" makes visual as well as statistical sense. The visual idea of OLS is that we are trying to find a way to get a straight line that goes through the data points by coming as close to possible to all of them on average.
2. The extended logic behind that is to make all data points on a scatterplot form a horizontal straight line by using a mathematical transformation of the *y*-axis into a function of the *x*-axis. That is basically what a model does from a visual perspective.
3. Finally, the logic of minimizing the distance between the line and the data points implies that the true result of a model is never the model on its own, but the combination of the line and the data points.

So you can see that a lot about statistical modelling can be translated into visual terms without deviating too much from canonical terminology. Let's take a look at R code to run linear regression models.

<!-- add Visually weighted regression -->

```{r vwreg1, eval=FALSE, echo=FALSE, results='hide'}
library(foreign)
source("code/vwreg.r")
```

```{r vwreg2, eval=FALSE, echo=FALSE, results='hide'}
# Import the data.
if (file.exists("data/qog_basic_cs.csv")) {
  # Open local copy.
  qog <- read.csv("data/qog_basic_cs.csv")
} else {
  print("Cannot find dataset, loading from web...")
  # Retrieve file.
  qog <- read.csv("http://www.qog.pol.gu.se/digitalAssets/1373/1373417_qog_basic_cs_csv_120608.csv", header = FALSE)
  # Save local copy.
  write.csv(qog, file = "data/qog_basic_cs.csv")
}
```

```{r vwreg3, eval=FALSE, echo=FALSE, message=FALSE, results='hide', cache=TRUE}
p1 <- vwReg(wdi_fr ~ bl_asyf25 + I(bl_asyf25^2), qog, method=lm)
```

```{r vwreg4, eval=FALSE, echo=FALSE}
p1 + ylab("Fertility rate") + xlab("Average education years among 25+ year-old females")
```

> __Next__: [Practice](083_practice.html)
