<style>@import url(style.css);</style>
[Introduction to Data Analysis](index.html "Course index")

# 10.1. Geocoded information

Most of the geocoding work that we want to show here can be [done with `ggmap`][milanor-ggmap], a map extension to `ggplot2` that uses Google Maps or other services to provide geocoded map backgrounds.

[milanor-ggmap]: http://www.milanor.net/blog/?p=534
[wb-aid]: http://aiddata.org/content/index/AidData-Raw/geocoded-data

```{r packages, message = FALSE}
packages <- c("downloader", "ggmap", "plyr")
packages <- lapply(packages, FUN = function(x) {
  if(!require(x, character.only = TRUE)) {
    install.packages(x)
    library(x, character.only = TRUE)
  }
})
```

Let's plot [World Bank projects in Africa][wb-aid].
 
```{r africa-data}
# Get the data.
url = "http://aiddata.org/weceem_uploads/_ROOT/File/geocoding/AllWorldBank_IBRDIDA.zip"
zip = "data/wb.projects.zip"
if(!file.exists(zip)) download(url, zip, mode = "wb")
# Read from the ZIP file.
wb = read.csv(unz(zip, "AllWorldBank_IBRDIDA.csv"))
# Subset to Africa.
wb = subset(wb, Region == "AFRICA")
# Inspect variables.
v = c("Project.ID", "Latitude", "Longitude", "Country", "Total.Amt")
head(wb)[v]
```

The next step involves injecting some information from an online map bank into R. The `get_map` function calls the [Open Street Map][osm] API, which returns a `ggmap` raster object that you can pass to `ggmap`. That map can be overplotted with geocododed information, as shown below with aid projects from the World Bank, colored by country and sized by total amount of funding.

```{r africa-osm-auto, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 9, tidy = FALSE}
# Get OpenStreetMap data.
map =  get_map(location = 'Africa', zoom = 4, source = "osm")
# Plot World Bank projects.
ggmap(map) + 
  geom_point(data = wb, 
             aes(x = Longitude, y = Latitude, color = Country, size = Total.Amt),
             alpha = .3) + 
  scale_size_area(max_size = 8) + 
  labs(y = NULL, x = NULL) +
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(),
        legend.position = "none")
```

The `get_map` function puts more than one map service at your fingertips: here's an example with the [Stamen watercolor][stamen] map. The map still uses a different color for each country and passes the same plot options to remove the unnecessary axis and title information. It would be easy to write a quick map function to avoir recoding each plot (have a try).

[stamen]: http://content.stamen.com/watercolor_process

```{r africa-stamen-auto, warning = FALSE, message = FALSE, fig.width = 12, fig.height = 9, tidy = FALSE, cache = TRUE}
# Get OpenStreetMap data.
ton = get_map(location = 'Africa', zoom = 4, source = "stamen", maptype = "watercolor")
# Plot World Bank projects.
ggmap(ton) + 
  geom_point(data = wb, 
             aes(x = Longitude, y = Latitude, color = Country, size = Total.Amt)) + 
  scale_size_area(max_size = 8) +
  labs(y = NULL, x = NULL) +
  theme(axis.text = element_blank(), 
        axis.ticks = element_blank(),
        legend.position = "none")
```

If you geodata consists of routes, like flights, it contains vertices of the form $(x_1, y_1) - (x_2,y_2)$ (start-end points) that can be plotted by latitude and longitude. If you need to visualize these ties, there are ways to [plot network data over maps][yau-circles] with a different R package. We will quickly return to network data next week, in a simpler context.

[yau-circles]: http://flowingdata.com/2011/05/11/how-to-map-connections-with-great-circles/

> __Next__: [Choropleth maps](102_choropleths.html).
