<style>@import url(style.css);</style>
[Introduction to Data Analysis](index.html "Course index")

# 10.1. Geocoded information

Most of the work can be [done with ggmap](http://www.milanor.net/blog/?p=534). Let's plot [airport traffic](http://www.milanor.net/blog/?p=594), using an online tutorial by Max Marchi. It shows how to plot geocoded airports and additional information on Google Maps, from R.

```{r packages, message = FALSE}
packages <- c("downloader", "ggmap", "plyr")
packages <- lapply(packages, FUN = function(x) {
  if(!require(x, character.only = TRUE)) {
    install.packages(x)
    library(x, character.only = TRUE)
  }
})
```

```{r airports-data}
# Prepare the data.
url = "http://openflights.svn.sourceforge.net/viewvc/openflights/openflights/data/airports.dat"
file = "data/openflight-airports.txt"
if(!file.exists(file)) download(url, file, mode = "wb")
airports <- read.csv(file, header = FALSE)
names(airports) <- c("ID", "name", "city", "country", "IATA_FAA", "ICAO", "lat", "lon", "altitude", "timezone", "DST")
str(airports)
# Prepare more data.
url = "http://openflights.svn.sourceforge.net/viewvc/openflights/openflights/data/routes.dat"
file = "data/openflight-routes.txt"
if(!file.exists(file)) download(url, file, mode = "wb")
routes <- read.csv(file, header = FALSE)
names(routes) <- c("airline", "airlineID", "sourceAirport", "sourceAirportID", "destinationAirport", "destinationAirportID", "codeshare", "stops", "equipment")
str(routes)
```

```{r map-europe}
# Create a simple map of Europe, using Google Maps.
map <- get_map(location = 'Europe', zoom = 4)
ggmap(map)

# Plot the data by passing coordinates.
ggmap(map) + 
  geom_point(data = airports, aes(x = lon, y = lat), 
             color = "orangered", alpha = .5)
```

```{r}
# Aggregate trips into number of departures and arrivals.
departures <- ddply(routes, .(sourceAirportID), "nrow")
names(departures)[2] <- "flights"
arrivals <- ddply(routes, .(destinationAirportID), "nrow")
names(arrivals)[2] <- "flights"

# Create datasets by matching data to airports.
airportD <- merge(airports, departures, by.x = "ID", by.y = "sourceAirportID")
airportA <- merge(airports, arrivals, by.x = "ID", by.y = "destinationAirportID")

# Plot airports proportionally to the square root of flights.
ggmap(map) +
  geom_point(data = airportD, aes(x = lon, y = lat, size = sqrt(flights)), alpha = .5)
```


```{r}
# Prepare a sequence to pass to square root legends.
seq <- c(1, 5, 10, 50, 100, 500)

# Adjust the legend.
ggmap(map) +
  geom_point(data = airportD, aes(x = lon, y = lat, size = sqrt(flights)), alpha = .5) +
  scale_size_area(breaks = sqrt(seq), labels = seq, name = "Departing routes")
```

```{r}
# Combine departures and arrivals in a single dataset.
airportD$type <- "Departures"
airportA$type <- "Arrivals"
airportDA <- rbind(airportD, airportA)

# Map the data.
ggmap(map) +
  geom_point(data = airportDA, aes(x = lon, y = lat, size = sqrt(flights)), alpha = .5)
```

```{r}
# Adjust the legend and separate by type of flight.
ggmap(map) +
  geom_point(data = airportDA, aes(x = lon, y = lat, size = sqrt(flights)), alpha = .5) +
  scale_size_area(breaks = sqrt(seq), labels = seq, name = "Routes") +
  facet_grid(. ~ type)
```

There are also ways to [plot network data over maps](http://flowingdata.com/2011/05/11/how-to-map-connections-with-great-circles/), if your data contains vertices of the form $(x_1, y_1) - (x_2,y_2)$ (start-end points) that can be plotted by latitude and longitude.

> __Next__: [Choropleth maps](102_choropleths.html).
