<style>@import url(style.css);</style>

# 1.1. R

[R](http://www.r-project.org/) is a software that uses a specific language to manipulate statistical objects. Researchers and statisticians are [increasingly](http://r4stats.com/articles/popularity/) adopting R as their software of choice, and the R user base goes well beyond academia. There is a lively R [blogging scene](http://www.r-bloggers.com/), tons of support on websites such as [StackOverflow](http://stackoverflow.com/questions/tagged/r), and many [R User Groups](http://www.r-bloggers.com/RUG/) exist on a [worldwide scale](http://blog.revolutionanalytics.com/local-r-groups.html).

The R language has its roots in the S language developed by AT&T, which also developed the C language. It is not the only possible language for statistical analysis: there are [quite a few software options](http://brenocon.com/blog/2009/02/comparison-of-data-analysis-packages-r-matlab-scipy-excel-sas-spss-stata/) for data analysis. R is as free as it is powerful, but it comes with [some drawbacks](http://badhessian.org/seven-reasons-to-use-r-for-social-network-analysis-and-three-reasons-against/#more-446), and the major disadvantage here is that its learning curve is pretty steep. Do not panic: we'll see as many examples as we need to make it work.

Install the latest version of R. You will need admin privileges on your computer to do so, and if you need help, turn to a video tutorial [for Mac](http://www.youtube.com/watch?v=ffG6jwM-8NQ) or [for Windows](http://www.youtube.com/watch?v=3zEbzZ9PqBo). Then open R and identify the __working directory__, which is R will look if you try to open or save stuff. This directory needs to point to course folder so that it can load the course material from there.

## Commands

The next step is to locate the console, where you will type commands and read their results. The following example shows how to store a text object and a numerical object, and how to print them as a vector that ends with a missing value, which are coded as `NA` in R:

```{r hello}
txt <- "Hello World!"
num <- (8^2)/5
print(c(txt,num,NA))
```

Let's go through all parts of this example. The basic `<-` operator performs an **allocation** of value `"Hello World!"` to the object `txt`. Text objects are called strings. The same operation is repeated with a numerical result to the `num` object. We then bind these objects together with the `c()` command that creates a **vector** with the `txt` and `num` values, followed by a missing value. The result is specifically shown on screen with the `print()` command, but this command can also be omitted.

In class, we will show how to use common mathematical functions in R, and how to manipulate simple objects like sequences, vectors and matrixes. Our practice session will be used to setup all laptops and to introduce the software interface.

## Packages
                                                                                                                                                                                                                                                                                                                                                                                                                 
Some particular commands need to be installed into R to work. This might be done through compressed archives of packages, or through the online [CRAN](http://cran.r-project.org/) central repository. This makes the software modular and has encouraged the development of packages for statistics and visualization by a wide audience of R programmers.

```{r default-packages}
# The default packages.
getOption("defaultPackages")
```

Let's install a few packages from a short list, and precede that with a comment on a separate line, which needs to start by `#` to be recognized as such by R:

```{r installing, eval=FALSE, include=TRUE}
# Install additional packages.
list <- c("ggplot2","knitr")
install.packages(list)
```

Once installed, the additional packages need to be loaded in memory with the `library()` command:

```{r loading, eval=FALSE, include=TRUE}
library(ggplot2)
library(knitr)
```

If we ever need to update these packages, we will simply type `update.packages()` and then agree to the updates that it suggests by comparing local versions of the packages with their online CRAN repository version.

> Next: [RStudio](12_rstudio.html).
