<style>@import url(style.css);</style>
<small>[Introduction to Data Analysis](index.html)</small>

# 1.3. Help pages

Type in `help.start()` to access the index of documentation pages in R. The documentation is only that helpful, so looking for help online is frequent practice. Finding help is a skill in itself, and there's an etiquette to [asking questions online](https://www.youtube.com/watch?v=ZFaWxxzouCY).

If you enjoy video tutorials, the [two-minute screencasts](http://www.twotorials.com/) by Anthony Damico are particularly recommendable as a help base: make sure to bookmark that page and get into the habit of collecting useful help resources that way.

## Help for commands

If you are looking for help on a specific command, like `setwd` (which sets the working directory), then type `?setwd`. If you are looking for a keyword, e.g. how to plot a heatmap, type `??heatmap` to perform a fuzzy search.

Here is how it looks like in practice:

    ?setwd
    ??heatmap

Finally, this command will bring up help for a whole package.

    help(package="ggplot2")

The help pages open in the bottom-right corner of RStudio; you can customize that in the application's preferences. You might want to try out `?library` and `?c` to learn about the commands used in previous examples. 

Documentation pages in R contain precise instructions on the syntax of each command: read the 'Description', 'Usage' and 'Arguments' sections first, then turn to 'Details' (not always needed) and always check out the 'Examples' section.

Do not hate the documentation! _Everyone_ finds it difficult to read, and there are good reasons why the technical documentation is written how it is. Turn to your handbooks, to tutorials and to the class material when the documentation does not help.

## Help for functions

You can type a command without its final brackets to see its internal workings:

```{r help-functions}
# Core mechanics of sum().
sum
# Core mechanics of median().
median
```

For a much more complex illustration, try this:

```{r help-functions-2, results='hide'}
# Core mechanics of ls().
ls
```

Admittedly, the output of doing this won't be that helpful in most cases.

## Practice is required

R is difficult. [Even professional programmers](http://www.johndcook.com/R_language_for_programmers.html) know that it takes some time to make sense out of it. The upside is, once you make sense out of R, it means that you have learnt a lot. A bit, perhaps, like learning German grammar or Spanish tenses.

Take some regular time with the software. Use R two hours per week, and fiddle with the examples given here: try out your own, even if it means modifying them just slightly. Do not neglect fishing for [other tutorials](03_links.html) online, there's tons of them.

> __Next week__: [Basic R operators](20_basics.html).
