<style>@import url(style.css);</style>
<small>[Introduction to Data Analysis](index.html)</small>

# 2.1. Command syntax

Before we go further into manipulating objects, here's a quick tour of R syntax. You will inevitably make mistakes when writing R commands, which is fine -- nothing (bad) happens when you do. To correct a mistake, simply press `UpArrow` (&uarr;) to go back into your last console input and fix it.

```{r broken-hello-world, eval=FALSE, tidy=FALSE}
# This will break. Try it.
print(Hello World!)
# Press UpArrow, select text, add double quotes.
# This will now run fine.
print("Hello World")
```

R syntax can be a bit strange. For some (probably good) reason, the `install.packages` command takes quotes, whereas the `library` command does not.

```{r quotes-strangeness, eval=FALSE}
# This needs quotes.
install.packages("ggplot2")
# This does not.
library(ggplot2)
# But it can.
library("ggplot2")
# There's also another (less noisy) way to load packages.
require(RCurl)
# Thankfully, it also accepts quotes.
require("RCurl")
# Oh, and single quotes can work too...
library('ggplot2')
# ... And that covers the whole spectrum of quotes strangeness.
require('RCurl')
```

Note that brackets are omnipresent in R syntax: every function (or command) will require them. The bits written inside the brackets and separated by commas are called arguments. The `cat` command, for instance, accepts any number of arguments.

```{r cat}
# Print some text, the full date and some text again, with no separators.
cat("Hello World! Today is ", date(), ".", sep = "")
```

A very important aspect of R syntax is *assignment*. To put some values into an object, you need two things, separated by a right-left arrow (or by a single equal sign if you prefer). The example below creates an object that holds my Body Mass Index, and then evaluates its normality.

```{r my-bmi}
# Compute my Body Mass Index.
bmi <- round(703*134 / (70^2), 1)
# Create a text object (called a string).
assessment <- "normal"
# Modify the assessment statement if BMI is below 18 or above 25.
if(bmi < 18) assessment <- "below normal"
if(bmi > 25) assessment <- "above normal"
cat("My BMI is approximately", bmi, ", which is", assessment)
```

In fact, it would make more sense to write a Body Mass Index *function*, which is not very difficult if you can survive the additional brackets. We will come back shortly to writing functions next week.

```{r bmi-function}
# A simple Body Mass Index function.
bmi <- function(weight, height, digits = 2) {
  round(weight*703 / (height^2), digits)
}
# An object called 'bmi' now appears in your Workspace.
# Check result.
bmi
# This object is a function.
class(bmi)
# Example, with default argument of 2 digits.
bmi(weight = 134, height = 70)
# Another example, this time with no digits.
bmi(weight = 134, height = 70, digits = 0)
```

As a very quick exercise, create the `bmi.quetelet` function that computes the BMI in its [original metric units](https://en.wikipedia.org/wiki/Body_mass_index), and compute your own figure. Here's the [code solution](https://raw.github.com/briatte/ida/master/code/21_bmi_quetelet.R), with some code to show how to plot a single variable function.

The syntax of each command requires a bit of memorization, and lots of practice. Nobody types functional code in R without going through cycles of trial-and-error with it. Programming requires some discipline, some logic and a lot of patience.

A final thing: it is good practice to write code in a certain style. For example, R is insensitive to extra blank space, but it is good practice to use blank space only when it helps to align elements. The [Google R Style Guide](https://google-styleguide.googlecode.com/svn/trunk/google-r-style.html) is a good starting point to learn a few common conventions.

Let's now turn to manipulating simple things in R.

> __Next__: [Vectors and matrices](22_vectors.html).
