<style>@import url(style.css);</style>
<small>[Introduction to Data Analysis](index.html)</small>

# 2.1. Command syntax

R is a programming environment in which you can submit instructions for the software to execute, or "run". To type a command, move your cursor to the Console with `Ctrl-1` in RStudio, type in some command, and press `Enter`. Do not worry if you get an error, it will not break anything.

```{r hello-world}
print("Hello World!")
```

R will generally print results without you asking through the `print` command. Notice how the command follows a strict syntax: the command above will not work if you omit the brackets or quotes, or if you make a typo in `print`. To correct a mistake, simply press `UpArrow` to go back into your console input and fix the issue.

There is often more than one way to get what you want from R:
  
```{r cat}
cat("Hello World! Today is",date())
```

__The syntax of each command requires a bit of memorization, and lots of practice.__ I have never seen anyone type straight working code in R: everyone goes through cycles of trial-and-error with it, because the R language is based on rather old languages like Fortran and S that make its syntax often counter-intuitive (more on that with loops later on).

We should start by learning to manipulate simple things in R.