<style>@import url(style.css);</style>
<small>[Introduction to Data Analysis](index.html)</small>

# 2.2. Vectors and matrices

In R, everything is a *vector*. If it's not a vector, it's a special kind of vector.

What is a vector? A vector is an ordered list of $k$ things. The first thing is item #1 of the vector, the second thing is item #2 and so on. Here's a sequence of $k = 4$ integer values:

```{r seq}
# A sequence.
1:4
```

We are going to store the sequence into an object called `x` and verify that it is indeed a vector. And of course it is. It's a vector of dimension 4.

```{r vector-check}
# Store it.
x <- 1:4
# Is it a vector?
is.vector(x)
```

The examples below are also vectors:

```{r vectors-everywhere}
# Also a vector.
y <- "This is a vector."
# Check.
is.vector(y)
# Also a vector.
z <- 1
# Check.
is.vector(z)
# What about...
m <- cbind(1:3,4:6)
```

Okay, that's a *matrix*. A vector is a $1 \times k$ unidimensional matrix. 

So in fact, in R, everything is a matrix, or some special kind of a matrix.

Welcome to R.

## What goes into a vector

Vectors are just lists of stuff. That stuff must be of the same type: you cannot mix numbers and text, for instance. If you do, R will convert all items of the vector to a single type. You can see the different types in the Workspace.

```{r vector-items}
# A vector of numbers.
x <- c(4, 6, 8)
# A vector of strings, i.e. characters, i.e. text.
y <- c("alea", "jacta", "est")
# A 'mixed' vector will be automatically converted.
z <- c(4, "a")
```

## A simple student grades example

Let's combine a fictional set of student grades into a vector called `grades`. Let's compute the mean, the median, and finally a whole bunch of summary statistics about the resulting object.

```{r grades}
# Create vector of random grades.
grades <- c(7, 13, 19, 8, 12)
# Check result.
grades
# Compute mean of grades vector.
mean(grades)
# Compute median of grades vector.
median(grades)
# Summarize grades vector.
summary(grades)
```

Note that R uses a precise taxonomy of objects, and that even though we just used the word "list" to describe a vector, a *list* is not the same as a *vector*. If you test the `grades` object with the `is.vector` command, the result will be `TRUE`, but it will be `FALSE` if you ask whether `grades` is a list.

```{r grades-object}
# Recall the grades object.
grades
# It's a vector.
is.vector(grades)
# Select its first element.
grades[1]
# Select more than one element by listing which values you want.
grades[1:3]
# Select a vector of values: 1, 2, 3 and 5.
grades[c(1:3,5)]
# As a student rightly noted, it makes more sense to ask this instead.
grades[-4]
# Vectors can be logical.
grades > 10
# Select a logical vector of values.
grades[grades >= 10]
```

What is shown here is important: in R, everything is an object. Here, `grades` is an object, just like the list of values that you want to show. This logic is everywhere in R. It is harder to learn than less flexible syntaxes, but it provides a lot of power to the user.

Let's see what a different object looks like. If we want to organize some data into a "row by column" structure, we need to use a matrix--a table of numerical values, if you prefer. The `as.matrix` command will turn `grades` into a matrix.

```{r grades-as-a-matrix}
# The grades object is not a matrix.
is.matrix(grades)
# If you make it into a matrix, the vector becomes a column of values.
as.matrix(grades)
```

Let's continue by adding a unique identifier to each row, in case we later shuffle the data. The `cbind` command will form a single matrix object out of several vector ones.

```{r grades-id}
# Show length of grades vector.
length(grades)
# Create a numbered sequence 1, 2, 3, ...
id <- c(1:length(grades))
# Check result.
id
# Add to the matrix under a new name.
grades.data <- cbind(id, grades)
# Check result.
grades.data
```

Now let's code if the student passes or fails the test. The `ifelse` command works pretty much like the `IF` command in spreadsheet editors. The `cbind` commands adds a column to the matrix, using the existing `grades` matrix and the newly created `pass` vector.

```{r grades-pass}
# Create a text object based on a logical condition.
pass <- ifelse(grades >= 10, "Pass", "Fail")
# Check result.
pass
# This is what it would look like when added to the grades data.
# Find the issue with that operation.
cbind(grades.data, pass)
# This solves the issue: convert the text column to numeric values.
cbind(grades.data, pass = factor(pass))
# Understand what happens when you factor a string (text) variable.
factor(pass)
# Rename grades vector.
grades1 <- grades
# Final conversion, with renaming of grades column.
grades.data <- cbind(id, grades1, pass = factor(pass))
# Check first five rows of result.
head(grades.data)
```

```{r grades-explore}
# First row, second cell.
grades.data[1,2]
# First row.
grades.data[1,]
# First two rows.
grades.data[1:2,]
# Third column.
grades.data[,3]
```

Let's finally add a column of random grades, compute the average for every row, and sort the data by ascending average.

```{r grades-mean}
# Create random vector of length 5.
grades2 <- round(20*runif(5),1)
# Check result.
head(grades2)
# Add to matrix.
grades.data <- cbind(grades.data, grades2)
# Check result.
grades.data
```

This ends our quick tour of how vectors and matrices work, but please make sure that you practice at home! You will need that practice to become familiar with R. And for the matrix-savvy, here's a [cheat sheet of matrix operations in R](https://gastonsanchez.wordpress.com/2012/08/30/cheat-sheet-of-matrix-operations-in-r/).

> __Next__: [Other R objects](23_objects.html).
