<style>@import url(style.css);</style>
<small>[Introduction to Data Analysis](index.html)</small>

# 2.2. Vectors and matrices

## A simple student grades example

Vectors are just lists of numerical values. Let's store a fictional set of student grades in memory under the name `grades`. The resulting object is a vector of values. Then, let's compute the mean, the median, and finally a whole bunch of summary statistics. Note that we will usually import the data rather than generate it manually in later exercises.

```{r grades}
# Create vector of random grades.
grades <- c(7.5,13,19,8,12)
# Check result.
grades
# Compute mean of grades vector.
mean(grades)
# Compute median of grades vector.
median(grades)
# Summarize grades vector.
summary(grades)
```

We will come back to summary statistics. Note that there is another way to create a variable and then check the result by enclosing the command in brackets.

```{r some-grades}
# Create vector of five elements.
(grades <- c(12, 9, 16, 4.5, 7))
# Save backup copy in object g.
g <- grades
```
Note that R uses a precise taxonomy of objects, and that even though we just used the word "list" to describe a vector, a *list* is not the same as a *vector*. If you test the `grades` object with the `is.vector` command, the result will be `TRUE`, but it will be `FALSE` if you ask whether `grades` is a list. You can turn `grades` into a list with the `as.list` command.

```{r grades-object}
# The grades object is a vector.
is.vector(grades)
# You can select any of its element.
grades[1]
# You can select more than one element by listing which values you want.
grades[1:3]
# You can provide a continuous sequence, as above, or a vector of values.
grades[c(1,3,5)]
```

What is shown here is important: in R, everything is an object. Here, `grades` is an object, just like the list of values that you want to show. This logic is everywhere in R. It is harder to learn than less flexible syntaxes, but it provides a lot of power to the user.

Let's see what a different object looks like. If we want to organize some data into a "row by column" structure, we need to use a matrix--a table of numerical values, if you prefer. The `as.matrix` command will turn `grades` into a matrix.

```{r grades-as-a-matrix}
# The grades object is not a matrix.
is.matrix(grades)
# If you make it into a matrix, the vector becomes a column of values.
as.matrix(grades)
```

We use the extra brackets around the command to print its results to the Console, in order to check that your commands are working as expected. In this example, we get to see what the `grades` object looks after becoming `grades_matrix`.

Let's continue by adding a unique identifier to each row, in case we later shuffle the data.

```{r grades-id}
# Show length of grades vector.
length(grades)
# Create a numbered sequence 1, 2, 3, ...
id <- c(1:length(grades))
# Check result.
id
# Add to the matrix under a new name.
grades.data <- cbind(id, grades)
# Check result.
grades.data
```

Now let's code if the student passes or fails the test. The `ifelse` command works pretty much like the `IF` command in spreadsheet editors. The `cbind` commands adds a column to the matrix, using the existing `grades` matrix and the newly created `pass` vector. We will routinely build datasets with such operations.

```{r grades-pass}
#
pass <- ifelse(grades >= 10, "Pass", "Fail")
#
cbind(grades.data, pass)
#
cbind(grades.data, pass = factor(pass))
#
grades.data <- cbind(grades.data,pass = factor(pass))
# Check result.
grades.data
```

Here, the `c(1:k)` command creates a sequence of numbers in the form of a vector of length `k`, where `k` is the number of rows in the `grades` matrix. The `seq(1:5)` command, or just `1:5`, would provide the same result.

```{r grades-explore}
grades.data[1,2]  # first row, second cell
grades.data[1,]   # first row
grades.data[1:2,] # first two rows
grades.data[,3]   # third column
```

Let's finally add a column of random grades, compute the average for every row, and sort the data by ascending average. This ends our quick tour of how vectors and matrices work (together), but the lab session for this week goes into more detail. *Please make sure that you practice at home on the weekly lab session!* You will need that practice to become familiar with R.

```{r grades-mean}
# Create random vector of length 5.
grades2 <- round(20*runif(5),1)
# Check result.
grades2
# Add to matrix.
grades.data <- cbind(grades.data, grades2)
# Check result.
grades.data
```

## More complex student grades

Let's use some real, anonymized data from Autumn 2012. These are the grades from my three mathematics classes in first year. I have removed any student identification, so you have to trust me that these are the real grades. We will use a package to create fake names in replacement.

```{r mygrades-import}
# Install randomNames package. Remember that R is case-sensitive.
install.packages("randomNames")
# Import the data from CSV format.
read.csv("data/grades.csv")
```

[Cheat sheet of matrix operations](https://gastonsanchez.wordpress.com/2012/08/30/cheat-sheet-of-matrix-operations-in-r/).

## Getting started with graphs

Let's produce [a few plots](http://flowingdata.com/2012/12/17/getting-started-with-charts-in-r/) from fake data vectors.

> __Next__: [Other R objects](23_objects.html).
