<style>@import url(style.css);</style>
<small>[Introduction to Data Analysis](index.html)</small>

# 3.1. Functions

Suppose that you find yourself doing the same computation over and over again. You do not want to type the whole calculations every time you need: you want to program a function that does the job for you.

Take a basic example: the `sum` command adds a vector of numbers together.

```{r sum}
# Create a vector of 99 random [0,1] values.
x <- runif(9)
# Check result: show first 5 values.
head(x)
# Add them up.
sum(x)
```

```{r add-function}
# Define function
add <- function(x,y) {
  x + y
}
# Example.
sum(2,4)
```

Functional programming focues on writing functions. Statistical computing courses like [the one taught by Cosma Shalizi](http://www.stat.cmu.edu/~cshalizi/statcomp/) contains tons of interesting examples of such functions. A simple one is shown below to illustrate the principle of recursion.

```{r crs-factorial}
# Calculate a factorial
# Input: a number (n)
# Output: the factorial of n
# Presumes: n is a single positive integer
my.factorial <- function(n) {
  if (n == 1) {  # Base case
    return(1)
  } else {  # Recursion
    return(n*my.factorial(n-1))
  }
}
```

This course might lead you to write simple functions, but for the moment, let's focus simply on plotting them with the `ggplot2` package, which makes it fairly easy. The example below shows the basic function $y = x$.

```{r fun-identity}
qplot(c(0,2), stat="function", fun=identity, geom="point")
```

Now for the most math-oriented, here's a sine function for your enjoyment.

```{r fun-sin}
qplot(c(-10,10), stat="function", fun=sin, geom="line")
```

And finally, a function that will be more useful to us: the exponential.

```{r fun-exp}
qplot(c(-10:10), stat="function", fun=exp, geom=c("line","point"))
```

> __Next__: [Iteration](32_iteration.html).
