<style>@import url(style.css);</style>
<small>[Introduction to Data Analysis](index.html)</small>

# 3.2. Loops

Suppose you want to remove a letter from a column of data. Now, suppose that you want to remove a letter from 9,000 columns of data. You are going to need a system to do the same thing over and over: you are going to need a loop.

## `for` and `while`

R understands `for` and `while` loops like many other programming languages. The idea is the same for both operators: allow to repeat an operation a certain number of times. The first example is as trivial as it gets.

```{r for}
# Set the counter.
n <- 3
# Loop n times.
for(i in 1:n) {
  cat("Hello!", i, "\n")
}
```

The next example shows how to build a sequence of integers through a `for` loop. Explain the output.

```{r r-for-vector}
# Set the counter.
n <- 9
# Create an empty vector.
sequence <- NULL
# Loop over 9 values.
for(i in 1:n) {
  # Add i to the sequence.
  sequence <- c(sequence, i)
  # Mean of the sequence.
  print(mean(sequence))
}
# Show final result.
sequence
```

A `while` loop operates almost on the same level.

```{r while}
# Set the counter.
i <- 5
# Decrease i by an increment of 1 until it reaches 0.
while(i > 0) {
  print(i <- i - 1)
}
```

## `apply` functions

Loops are fantastic but can become memory-intensive. R also offers [`apply` functions](http://badhessian.org/loops-matrices-and-apply-functions/) that can usually do what you might want to do with a loop, but we will keep that for later, when we look at data manipulation.

> __Next__: [Probability](33_proba.html).
