<style>@import url(style.css);</style>

# Descriptive statistics

## Extracting a mean over several observations

Let's use the Daily Kos dataset holding Obama, McCain and Romney scores for the 2008 and 2012 elections. The dataset holds data for each Congressional district.

```{r data-table1, message=FALSE}
# Load the data from file (created in Session 4.1).
dk_data <- read.csv("data/dailykos.csv")
# Review columns 1-5.
head(dk_data)[1:5]
```

We'll extract the state by creating a `state` variable from the first two letters of the `CD` variable.

```{r data-table2}
# Create a variable holding states.
dk_data$state <- substr(dk_data$CD,1,2)
# Check result: the new variable is at the end.
str(dk_data)
```

Let's now turn the dataset (which is a `data.frame` object) into a `data.table` object (without the missing values), and show the state-level average electoral support for Barack Obama in 2012.

```{r data-table3}
# Load data.table package.
library(data.table)
# Convert nonmissing data to a data.table object.
dk_table <- data.table(na.omit(dk_data))
# Check result.
head(dk_table)
# Compute mean Obama score by state.
dk_state <- dk_table[, mean(Obama.2012), by=state]
# Check result.
head(dk_state)
```

Before plotting the result, we sort the states by electoral support for Obama (`V1`). This works because a `data.table` works like a `data.frame` object.

```{r data-table4}
# Sort the data.
dk_state$state <- reorder(dk_state$state, dk_state$V1)
# Plot by decreasing electoral support.
p <- ggplot(dk_state, aes(V1, state)) + geom_point() + 
  geom_vline(xintercept = 50, colour="darkgray", linetype = "longdash") +
  labs(x = "% of popular vote for Obama in 2012", y = "State"); p
```

Let's finish by coloring state by majority party.

```{r data-table5}
# Create a red/blue variable.
dk_state$color <- ifelse(dk_state$V1 > 50, "blue", "red")
# Add a new layer of colored dots.
p + geom_point(colour=dk_state$color)
```
