<style>@import url(style.css);</style>

# Introduction to Data Analysis

This course is an introduction to analyzing data with the [R](http://www.r-project.org/) software. 
It features some mathematics and statistics as well as some statistical computing and data visualization. 
You will need a laptop with an Internet connection to follow the class.

```{r prepare-folder, include=FALSE, message=FALSE}
# course directory
setwd("~/Documents/Teaching/IDA/")
# backup code
zip(paste("admin/backup", Sys.Date(), "zip", sep = "."), dir(pattern = ".Rmd"))
# clean up
rm(list = ls())
```

```{r run-course, include=FALSE, message=FALSE}
require(foreach)
require(knitr)
# get files
all <- dir(pattern="[0-9]+(.*).Rmd")[1:2] # sequence is 2 + session number * 3 
# knitr setup
opts_chunk$set(fig.path = "plots/", comment = NA)
# go go go
lapply(all, FUN = function(x) {
  print(x)
  if(substr(x, 0, 2) != "00")
    purl(x, documentation = 0, output = paste0("code/", gsub("Rmd", "R", x)))
  knit2html(x)
})
# get HTML
html <- dir(pattern="(index|[0-9]{3,}_\\w+).html|style.css")
# get code
code <- dir("code", pattern="([0-9]{3,}_\\w+).R$")
# set path
path <- "~/Documents/Code/Websites/briatte.github.com/teaching/ida"
# copy to website
file.copy(html, path, overwrite = TRUE)
if(length(code) > 0) file.copy(paste0("code/", code), path, overwrite = TRUE)
# clean up
file.remove(gsub("Rmd", "md", all))
file.remove(gsub("Rmd", "html", all))
```

- [Introduction][001]
- [Readings][002]

[001]: 001_intro.html
[002]: 002_readings.html

> __Part 1: Introduction to Statistical Computing__. The course starts by one month dedicated to setting up R and learning its basic functionalities. All course logistics will be discussed in these weeks.

## 1. [Setup][010]

__Readings:__ Kabacoff, ch. 1 and Teetor, ch. 1 and 3.

- [Installing R][011]
- [Installing RStudio][012]
- [Practice][013]

[010]: 010_setup.html
[011]: 011_r.html
[012]: 012_rstudio.html
[013]: 013_practice.html

* * *

## 2. [Objects][020]

__Readings:__ Teetor, ch. 2 and 5.

- [Vectors and matrices][021]
- [Variables and factors][022]

[020]: 020_objects.html
[021]: 021_vectors.html
[022]: 022_variables.html

* * *

## 3. [Functions][030]

__Readings:__ Kabacoff, ch. 5, and Teetor, ch. 8.

<!-- Matloff, ch. 8. -->

- [Boolean logic][031]
- [Iteration and probability][032]

[030]: 030_functions.html
[031]: 031_logic.html
[032]: 032_iteration.html

* * *

## 4. [Data][040]

__Readings:__ Kabacoff, ch. 4, and Teetor, ch. 4 and 6-7.

- [Imports, exports, scraping][041]
- [Reshapes and aggregates][042]

[040]: 040_data.html
[041]: 041_dataio.html
[042]: 042_reshaping.html

* * *

> __Part 2: Introduction to Statistical Analysis__. The course continues by showcasing some statistical techniques, from finding clusters of related data to modelling relationships between several variables.

## 5. [Clusters](50_clusters.html)

__Readings:__ Kabacoff, ch. 14, and Teetor, ch. 13.4, 13.6 and 13.9.

- [Heatmaps](51_heatmaps.html)
- [Principal components](52_pca.html)
- [*k*-means](53_kmeans.html) ##merge with PCA

* * *

## 6. [Distributions](60_distributions.html)

__Readings:__ Kabacoff, ch. 6, and Teetor, ch. 10.

- [Descriptive statistics](61_descr.html) ##edit: name?
- [Probability density functions](62_pdf.html) ##edit: merge with below
- [Density functions](63_density.html) # ECDF, histograms, PDF (normal distribution)

* * *

## 7. [Differences](70_hyptests.html) ##edit: rename

__Readings:__ Kabacoff, ch. 7, and Teetor, ch. 9.

- [Confidence intervals](71_ci.html) ##edit: merge with t-test
- [Comparison of means](72_ttest.html)
- [Comparison of proportions](73_prtest.html)

* * *

## 8. [Models](80_lin.html) ##edit: rename

__Readings:__ Kabacoff, ch. 8 and 11, and Teetor, ch. 11. Skim ANOVA to focus on OLS.

- [Correlation and ordinary least squares](81_scatterplots.html) ##edit: rename
- [Linear regression](82_regression.html) ##edit: rename
- [Visually weighted regression](83_vwreg.html) ##edit: merge with linear reg.

* * *

> __Part 3: Introduction to Data Visualization__. The course ends by focusing on the graphic dimension of quantitative data. We will also try to have guest speakers to talk about their professional use of data.

## 9. [Visualization in time: Time series](90_ts.html)

__Reading:__ Teetor, ch. 14. # detrending, ARIMA

- [Autocorrelation](91_lags.html) ##edit
- [Smoothing](92_smoothing.html) ##edit
- ARIMA (perhaps) ##edit

* * *

## 10. [Visualization in space: Maps](100_maps.html)

__Reading:__ B. Sue Bell *et al.*, "[Current practices in spatial analysis of cancer data mapping health statistics to inform policymakers and the public](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1647272/)", 2006.

- [Choropleth maps](101_choropleth.html) ##edit
- [Google-powered maps](102_gmaps.html) ##edit
- Spatial dependence (perhaps) ##edit

* * *

## 11. [Networks](110_networks.html)

__Reading:__ David Siegel, "[Social Networks in Comparative Perspective](http://opensiuc.lib.siu.edu/cgi/viewcontent.cgi?article=1045&context=pn_wp)", 2011.

- [Measures of influence](111_influence.html) ##edit
- [Mining an online network](112_twitter.html) ##edit
- Plotting interactive networks (perhaps) ##edit

* * *

## 12. [Data-driven advances](120_data.html)

__Special session__ with guests (if everything goes well):

- [What is data-driven journalism?] <!-- with Alexandre Léchenet -->
- What is open data/government? <!-- with OKFN -->
- What is reproducible science? <!-- with Joël Gombin -->

* * *

We're done!

__Thanks to__ [Baptiste Coulmont](http://coulmont.com/), [Joël Gombin ](http://joelgombin.fr/) and [Timothée Poisot](https://github.com/tpoisot/RCourse2012) for very valuable advice and comments, to [GitHub](http://github.com/) for hosting and to users at [StackExchange](http://stackexchange.com/) for coding assistance.

Special thanks to the Sciences Po Reims staff and students for indefectible support.

__Inspired by__ [Christopher Adolph](http://faculty.washington.edu/cadolph/index.php?page=22), [Dave Armstrong](http://quantoid.net/ICPSRR.html), [Christopher Gangrud](https://github.com/christophergandrud/Introduction_to_Statistics_and_Data_Analysis_Yonsei/), [Andrew Gelman](http://andrewgelman.com/2012/12/textbook-for-data-visualization/), [Rebecca Nugent](http://www.stat.cmu.edu/~rnugent/teaching/introR/), [Gaston Sanchez](https://gastonsanchez.wordpress.com/), [David Sparks](http://is-r.tumblr.com/), [Hadley Wickham](http://had.co.nz/ggplot2/).

This course has its own [GitHub repository](https://github.com/briatte/ida); fork at will. This HTML version was compiled from source on `r format(Sys.time(), "%A %B %d, %Y")`.
