<style>@import url(style.css);</style>

# Introduction to Data Analysis

This course is an introduction to analyzing data with the [R](http://www.r-project.org/) software. It features some mathematics and statistics as well as some statistical computing and data visualization. You will need a laptop with an Internet connection to follow the class.

```{r prepare-folder, include=FALSE, message=FALSE}
# course directory
setwd("/Users/fr/Documents/Teaching/IDA/")
# backup code
zip(paste("../IDA-backup", Sys.Date(), "zip", sep = "."), dir(pattern = ".Rmd"))
# clean up
rm(list = ls())
file.remove(dir(pattern="[0-9]+(.*)\\.html"))
file.remove(dir(pattern="[0-9]+(.*)\\.md"))
```

```{r run-course, include=FALSE, message=FALSE}
require(foreach)
require(knitr)
# get files
all <- dir(pattern="[0-9]+(.*).Rmd")[c(1:4, 8, 12:30)] # sessions 0, 1, 2, 3, 4, 5, 12
# knitr setup
opts_chunk$set(fig.path = "plots/", comment = NA)
# go go go
foreach(i=1:length(all)) %do% {
  # R scripts
  purl(all[i], documentation = 0, output = paste0("code/", gsub("Rmd", "R", all[i])))
  # HTML pages
  knit2html(all[i])
  # Remove .md
  file.remove(gsub("Rmd", "md", all[i]))
}
```

```{r populate-website, include=FALSE, message=FALSE}
# get HTML
html <- dir(pattern="(index|[0-9]{2,}_\\w+).html|style.css")
# get code
code <- dir("code", pattern="([0-9]{2,}_\\w+).R$")
# set path
path <- "~/Documents/Code/Websites/briatte.github.com/teaching/ida"
# copy to website
file.copy(html, path, overwrite = TRUE)
file.copy(paste0("code/",code), path, overwrite = TRUE)
```

- [Introduction](01_intro.html)
- [Readings](02_readings.html)
- [Links](03_links.html)

> __Part 1: Introduction to Statistical Computing__. The course starts by one month dedicated to setting up R and learning its basic functionalities. All course logistics will be discussed in these weeks.

## 1. [Setup](10_setup.html)

__Readings:__ Kabacoff, ch. 1 and Teetor, ch. 1 and 3.

- [Installing R](11_r.html)
- [Installing RStudio](12_rstudio.html)
- [Help pages](13_help.html)

* * *

## 2. [Basic R operators](20_basics.html)

__Readings:__ Teetor, ch. 2 and 5.

- [Command syntax](21_syntax.html)
- [Vectors and matrices](22_vectors.html)
- [Other R objects](23_objects.html)

* * *

## 3. [Basic math operators](30_math.html)

__Readings:__ Kabacoff, ch. 5, and Teetor, ch. 8.

<!-- Matloff, ch. 8. -->

- [Functions](31_functions.html)
- [Iteration](32_iteration.html)
- [Probability](33_proba.html)

* * *

## 4. [Basic data operators](40_data.html)

__Readings:__ Kabacoff, ch. 4, and Teetor, ch. 4 and 6.

<!-- additional manipulations with strings and dates: Teetor, ch. 7 -->

- [Import/Export](41_dataio.html)
- [Reshaping](42_reshaping.html)
- [Scraping](43_scraping.html)

* * *

> __Part 2: Introduction to Statistical Analysis__. The course continues by showcasing some statistical techniques, from finding clusters of related data to modelling relationships between several variables.

## 5. [Clusters](50_clusters.html)

__Readings:__ Kabacoff, ch. 14, and Teetor, ch. 13.4, 13.6 and 13.9.

- [Heatmaps](51_heatmaps.html)
- [Principal components](52_pca.html)
- [*k*-means](53_kmeans.html)

* * *

## 6. [Distributions](60_distributions.html)

__Readings:__ Kabacoff, ch. 6, and Teetor, ch. 10.

- [Descriptive statistics](61_descr.html)
- [Probability density functions](62_pdf.html)
- [Empirical cumulative density functions](63_ecdf.html)

* * *

## 7. [Hypothesis tests](70_hyptests.html)

__Readings:__ Kabacoff, ch. 7, and Teetor, ch. 9.

- [Confidence intervals](71_ci.html)
- [Comparison of means](72_ttest.html)
- [Comparison of proportions](73_prtest.html)

* * *

## 8. [Linear models](80_lin.html)

__Readings:__ Kabacoff, ch. 8 and 11, and Teetor, ch. 11. Skim ANOVA to focus on OLS.

- [Correlation and scatterplots](81_scatterplots.html)
- [Linear regression](82_regression.html)
- [Visually weighted regression](83_vwreg.html)

* * *

> __Part 3: Introduction to Data Visualization__. The course ends by focusing on the graphic dimension of quantitative data. We will also try to have guest speakers to talk about their professional use of data.

## 9. [Visualization in time: Time series](90_ts.html)

__Reading:__ Teetor, ch. 14. We'll try to add ARIMA models to the course content, as they are highly relevant for finance students.

- [Autocorrelation](91_lags.html)
- [Smoothing](92_smoothing.html)
- ARIMA (perhaps)

* * *

## 10. [Visualization in space: Maps](100_maps.html)

__Reading:__ B. Sue Bell *et al.*, "[Current practices in spatial analysis of cancer data mapping health statistics to inform policymakers and the public](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1647272/)", 2006.

- [Choropleth maps](101_choropleth.html)
- [Google-powered maps](102_gmaps.html)
- Spatial dependence (perhaps)

* * *

## 11. [Networks](110_networks.html)

__Reading:__ David Siegel, "[Social Networks in Comparative Perspective](http://opensiuc.lib.siu.edu/cgi/viewcontent.cgi?article=1045&context=pn_wp)", 2011.

- [Measures of influence](111_influence.html)
- [Mining an online network](112_twitter.html)
- Plotting interactive networks (perhaps)

* * *

## 12. [Data-driven advances](120_data.html)

__Special session__ with guests (if everything goes well):

- [What is data-driven journalism?] <!-- with Alexandre Léchenet -->
- What is open data/government? <!-- with OKFN -->
- What is reproducible science? <!-- with Joël Gombin -->

* * *

We're done!

__Thanks to__ [Baptiste Coulmont](http://coulmont.com/), [Joël Gombin ](http://joelgombin.fr/) and [Timothée Poisot](https://github.com/tpoisot/RCourse2012) for very valuable advice and comments, to [GitHub](http://github.com/) for hosting and to users at [StackExchange](http://stackexchange.com/) for coding assistance.

Special thanks to the Sciences Po Reims staff and students for indefectible support.

__Inspired by__ [Christopher Adolph](http://faculty.washington.edu/cadolph/index.php?page=22), [Dave Armstrong](http://quantoid.net/ICPSRR.html), [Christopher Gangrud](https://github.com/christophergandrud/Introduction_to_Statistics_and_Data_Analysis_Yonsei/), [Andrew Gelman](http://andrewgelman.com/2012/12/textbook-for-data-visualization/), [Rebecca Nugent](http://www.stat.cmu.edu/~rnugent/teaching/introR/), [Gaston Sanchez](https://gastonsanchez.wordpress.com/), [David Sparks](http://is-r.tumblr.com/), [Hadley Wickham](http://had.co.nz/ggplot2/).

This course has its own [GitHub repository](https://github.com/briatte/ida); fork at will. This HTML version was compiled from source on `r format(Sys.time(), "%A %B %d, %Y")`.
