<style>@import url(style.css);</style>

# Introduction to Data Analysis

This course is an introduction to analyzing data with the [R](http://www.r-project.org/) software. 
It features some mathematics and statistics as well as some statistical computing and data visualization. 
You will need a laptop with an Internet connection to follow the class.

```{r prepare-folder, include=FALSE, message=FALSE}
# course directory
setwd("~/Documents/Teaching/IDA/")
# backup code
zip(paste("admin/backup", Sys.Date(), "zip", sep = "."), dir(pattern = ".Rmd"))
# clean up
rm(list = ls())
file.remove(dir(pattern="[0-9]{3,}_\\w{1,}\\.html$"))
file.remove(dir(pattern="[0-9]{3,}_\\w{1,}\\.md$"))
```

```{r run-course, include=FALSE, message=FALSE}
require(knitr)
# get files
all <- dir(pattern="[0-9]+(.*).Rmd")
all <- all[1:34] # sequence is 2 + session number * 4 
# knitr setup
opts_chunk$set(fig.path = "plots/", comment = NA)
# go go go
lapply(all, FUN = function(x) {
  if(substr(x, 0, 2) != "00")
    purl(x, documentation = 0, output = paste0("code/", gsub("Rmd", "R", x)))
  knit2html(x)
})
# get HTML
html <- dir(pattern="(index|[0-9]{3,}_\\w{1,}).html|style.css")
# get code
code <- dir("code", pattern="([0-9]{1,}_\\w+).R$")
# set path
path <- "~/Documents/Code/Websites/briatte.github.com/teaching/ida"
# copy to website
file.copy(html, path, overwrite = TRUE)
if(length(code) > 0) file.copy(paste0("code/", code), path, overwrite = TRUE)
# clean up
file.remove(gsub("Rmd", "md", all))
file.remove(gsub("Rmd", "html", all))
```

- [Introduction][001]
- [Readings][002]

[001]: 001_intro.html
[002]: 002_readings.html

> __Part 1: Introduction to Statistical Computing__. The course starts by one month dedicated to setting up R and learning its basic functionalities. All course logistics will be discussed in these weeks.

## 1. [Setup][010]

__Readings:__ Kabacoff, ch. 1 and Teetor, ch. 1 and 3.

- [Installing R][011]
- [Installing RStudio][012]
- [Practice][013]

[010]: 010_setup.html
[011]: 011_r.html
[012]: 012_rstudio.html
[013]: 013_practice.html

* * *

## 2. [Objects][020]

__Readings:__ Teetor, ch. 2 and 5.

- [Vectors and matrices][021]
- [Variables and factors][022]
- [Practice][023]

[020]: 020_objects.html
[021]: 021_vectors.html
[022]: 022_variables.html
[023]: 023_practice.html

* * *

## 3. [Functions][030]

__Readings:__ Kabacoff, ch. 5, and Teetor, ch. 8.

<!-- Matloff, ch. 8. -->

- [Conditionals][031]
- [Iteration][032]
- [Practice][033]

[030]: 030_functions.html
[031]: 031_conditionals.html
[032]: 032_iteration.html
[033]: 033_practice.html

* * *

## 4. [Data][040]

__Readings:__ Kabacoff, ch. 4, and Teetor, ch. 4 and 6-7.

- [Imports and exports][041]
- [Reshapes and aggregates][042]
- [Practice][043]

[040]: 040_data.html
[041]: 041_dataio.html
[042]: 042_reshaping.html
[043]: 043_practice.html

* * *

> __Part 2: Introduction to Statistical Analysis__. The course continues by showcasing some statistical techniques, from finding clusters of related data to modelling relationships between several variables.

## 5. [Clusters][050]

__Readings:__ Kabacoff, ch. 14, and Teetor, ch. 13.4, 13.6 and 13.9.

- [Heatmaps][051]
- [Principal components][052]
- [Practice][053]

[050]: 050_clusters.html
[051]: 051_heatmaps.html
[052]: 052_pca.html
[053]: 053_practice.html

* * *

## 6. [Distributions][060]

__Readings:__ Kabacoff, ch. 6, Teetor, ch. 10, and Urdan, ch. 4-6. See also Urdan, ch. 1-3, if you have forgotten everything about statistics.

- [Descriptive statistics][061]
- [Density functions][062]
- [Practice][063]

[060]: 60_distributions.html
[061]: 061_description.html
[062]: 062_density.html
[063]: 063_practice.html

* * *

## 7. [Differences][070]

__Readings:__ Kabacoff, ch. 7, Teetor, ch. 9, and Urdan, ch. 7, 9 and 14.

- [Confidence intervals][071]
- [Significance tests][072]
- [Practice][073]

[070]: 070_differences.html
[071]: 071_intervals.html
[072]: 072_tests.html
[073]: 073_practice.html

* * *

## 8. [Models][080]

__Readings:__ Kabacoff, ch. 8 and 11, Teetor, ch. 11, and Urdan, ch. 8, 10 and 13. Skim ANOVA to focus on OLS (simple and multiple linear regression).

- [Linear correlation][081]
- [Ordinary least squares][082]
- [Practice][083]

[080]: 080_models.html
[081]: 081_correlation.html
[082]: 082_ols.html
[083]: 083_practice.html

* * *

> __Part 3: Introduction to Data Visualization__. The course ends by focusing on the graphic dimension of quantitative data. We will also try to have guest speakers to talk about their professional use of data.

## 9. [Visualization in time: Time series][090]

__Reading:__ Teetor, ch. 14. Focus on detrending and read ARIMA only if you plan to earn millions as a financial analyst.

- [Autocorrelation][091]
- [Smoothing][092]
- [Practice][093]

[090]: 090_ts.html
[091]: 091_lags.html
[092]: 092_smoothing.html
[093]: 093_practice.html

* * *

## 10. [Visualization in space: Maps][100]

__Reading:__ B. Sue Bell *et al.*, "[Current practices in spatial analysis of cancer data mapping health statistics to inform policymakers and the public](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1647272/)", 2006.

- [Choropleth maps][101]
- [Spatial dependence][102]
- **Special guest**: [Joël Gombin][jg] on open science

[100]: 100_maps.html
[101]: 101_choropleths.html
[102]: 102_spatialdep.html
[jg]: http://www.joelgombin.fr

* * *

## 11. [Networks][110]

__Reading:__ David Siegel, "[Social Networks in Comparative Perspective](http://opensiuc.lib.siu.edu/cgi/viewcontent.cgi?article=1045&context=pn_wp)", 2011.

- [Measures of influence][111]
- [Mining an online network][112]
- **Special guest**: [Alexandre Léchenet][al] on data journalism

[110]: 110_networks.html
[111]: 111_influence.html
[112]: 112_twitter.html
[al]: http://alphoenix.net


* * *

## 12. [Data-driven advances][120]

__Final session__: the last special guest [Samuel Goëta][sg], who will talk about open government in the [Distinguished Lecture Series][dls-talk].

- What is data-driven journalism?
- What is open data/government?
- What is reproducible science?

[120]: 120_data.html
[sg]: http://coulisses-opendata.com
[dls-talk]: https://www.facebook.com/events/123760981144191/

* * *

We're done!

__Thanks to__ [Baptiste Coulmont](http://coulmont.com/), [Joël Gombin ](http://joelgombin.fr/) and [Timothée Poisot](https://github.com/tpoisot/RCourse2012) for very valuable advice and comments, to [GitHub](http://github.com/) for hosting and to users at [StackExchange](http://stackexchange.com/) for coding assistance.

Special thanks to the Sciences Po Reims staff and students for indefectible support.

__Inspired by__ [Christopher Adolph](http://faculty.washington.edu/cadolph/index.php?page=22), [Dave Armstrong](http://quantoid.net/ICPSRR.html), [Christopher Gangrud](https://github.com/christophergandrud/Introduction_to_Statistics_and_Data_Analysis_Yonsei/), [Andrew Gelman](http://andrewgelman.com/2012/12/textbook-for-data-visualization/), [Rebecca Nugent](http://www.stat.cmu.edu/~rnugent/teaching/introR/), [Gaston Sanchez](https://gastonsanchez.wordpress.com/), [David Sparks](http://is-r.tumblr.com/), [Hadley Wickham](http://had.co.nz/ggplot2/).

This course has its own [GitHub repository](https://github.com/briatte/ida); fork at will. This HTML version was compiled from source on `r format(Sys.time(), "%A %B %d, %Y")`.
